Type theories as quotient inductive-recursive types

### 2025-02-07 tex/main.tex
Some revisions after the meeting 1. definitional to strict 2. comment out heterogeneous identity;

\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate,final]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"

@@

\begin{abstract}
  Defining type theory in type theory is a long-standing challenge and a benchmark for the strength and expressivity of type theory and a proof assistant.
  Altenkirch and Kaposi used quotient inductive-inductive types to represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate substitution calculus and type theory with a universe and $\Pi$-types, and compare ours with quotient inductive-inductive definitions.
\end{abstract}

@@

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
\LT{terminology: strict vs.\ definitional vs.\ judgemental?}
%

%However, we found that using quotient inductive-inductive-recursive types alone were not sufficient for a smooth development.
%%
%%Not only do we want the action of substitutions on types to compute definitionally, but we also want as many equations to hold strictly as possible:
%%
%%on the one hand, we want substitutions to distribute over type formers, but on the other hand we want substitutions to satisfy composition and unit laws of the category of context and substitutions up to definitional equality.
%%
%Instead, we also needed to make use of functions defined by overlapping patterns.
%%
%In a proof assistant such as Agda, this can be emulated by Agda's experimental support for rewrite rules, but with an additional proof burden on the user to show (informally) that their use is justified meta-theoretically.
%%
%To get some help with this task by Agda's confluence checker, it is often necessary to needlessly complicate a definition to convince the checker that the proposed rewrite rules are indeed confluent.

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B(x)$.
%
We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.

We write $x \jdeq y$ for definitional equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P(x)$, $u : P(y)$ and $p : x =^A y$. One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$, where $\transfibf{P} : (p : x =^A y) \to P(x) \to P(y)$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
%
Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.

@@

Since $\Ty$ is inductively defined, we add a base case to ensure that $\Ty\;\Gamma$ is not empty for every $\Gamma$:
%
%If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
%For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
%Hence another constructor is introduced:
\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
\end{alignat*}
For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ for some $p: [\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\tau]\,A$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

@@

To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution does not compute since it is a constructor, leading to type mismatches occurring with substitution.

@@

However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.

@@

%where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 

@@

where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the definitional equality $[\sigma]\;\UU \equiv \UU$.

@@

%That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

To ensure that type substitution is definitional for all %\emph
{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

We revisit \Cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
\begin{example}
  The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is definitionally equal to $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$, i.e.\ the type of the term on the RHS.
  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are definitionally equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
  Consequently, the back-of-the-envelope calculation~\eqref{eq:pi2-comp-proof} is \emph{formally correct}.
\end{example}

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
\LT[noinline]{Should we mention any general condition for coherence?}
\Fredrik[noinline]{Is there one?}
\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

  We show the remaining cases by induction on $A$.
  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of the identities with $A \defeq \UU$ reduce to $\UU$.
\end{proof}
\begin{remark}\label{re:coherence-proof}
  Note that in the above proof the first four identities hold definitionally by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three identities follow from \eqref{eq:def-type-subst-5} by induction on $A$.
  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how congruence rules for type substitution is defined.
\end{remark}

\LT{We may wonder if the last three identities can be made definitional.
Can we find a complete set of rules?}

@@

In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes definitional. 
Likewise, the equality constructor $[]\El$ may be replaced by 
\begin{alignat*}{3}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u})
\end{alignat*}
Yet, this naive change breaks the local confluence!
The problem is that there is now additional interaction between types and terms. Observe that the local peak
\[
  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u).
\]
cannot reduce to the same term, since the term substitution $[\idS]\,u$ is merely a constructor.

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
\end{proof}

@@

For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are definitional.
we do not need any transport since $[;]_T$ and $[]\Pi$ are definitional.
However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositionally but not definitionally equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.

>>>>

\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"

@@

\begin{abstract}
  Defining type theory in type theory is a long-standing challenge and a benchmark for the strength and expressivity of type theory and a proof assistant.
  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our definition with quotient inductive-inductive definitions.
\end{abstract}

@@

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
%

%However, we found that using quotient inductive-inductive-recursive types alone were not sufficient for a smooth development.
%%
%%Not only do we want the action of substitutions on types to compute strictly, but we also want as many equations to hold strictly as possible:
%%
%%on the one hand, we want substitutions to distribute over type formers, but on the other hand we want substitutions to satisfy composition and unit laws of the category of context and substitutions up to strict equality.
%%
%Instead, we also needed to make use of functions defined by overlapping patterns.
%%
%In a proof assistant such as Agda, this can be emulated by Agda's experimental support for rewrite rules, but with an additional proof burden on the user to show (informally) that their use is justified meta-theoretically.
%%
%To get some help with this task by Agda's confluence checker, it is often necessary to needlessly complicate a definition to convince the checker that the proposed rewrite rules are indeed confluent.

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B(x)$.
%
We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.

We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P(x)$, $u : P(y)$ and $p : x =^A y$. One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$, where $\transfibf{P} : (p : x =^A y) \to P(x) \to P(y)$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
%
%Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
%As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.

@@

Since $\Ty$ is inductively defined, we add a base case $\UU : \implicit{\Gamma} \Ty\,\Gamma$ to ensure that $\Ty\;\Gamma$ is not empty.
%for every $\Gamma$:
%
%If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
%For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
%Hence another constructor is introduced:
%\begin{alignat*}{3}
%  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
%\end{alignat*}
For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following identity apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ for some $p: [\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\tau]\,A$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

@@

To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules strict, since the root cause is that type substitution does not compute since it is a constructor, leading to type mismatches occurring with substitution.

@@

However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.

@@

%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 

@@

where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.

@@

%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

To ensure that type substitution is strict for all %\emph
{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

We revisit \cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
\begin{example}
  The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is strictly equal to the type $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$ of the term on the RHS.
  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are strictly equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
  Consequently, the back-of-the-envelope calculation~\eqref{eq:pi2-comp-proof} is \emph{formally correct}.
\end{example}

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
\LT[noinline]{Should we mention any general condition for coherence?}
\Fredrik[noinline]{Is there one?}
\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become strict, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

  We show the remaining cases by induction on $A$.
  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of the identities with $A \defeq \UU$ reduce to $\UU$.
\end{proof}
\begin{remark}\label{re:coherence-proof}
  Note that in the above proof the first four identities hold strictly by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three identities follow from \eqref{eq:def-type-subst-5} by induction on $A$.
  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how congruence rules for type substitution is defined.
\end{remark}

\LT{We may wonder if the last three identities can be made strict.
Can we find a complete set of rules?}

@@

In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by 
\begin{alignat*}{3}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u})
\end{alignat*}
Yet, this naive change breaks the local confluence!
The problem is that there is now additional interaction between types and terms. Observe that the local peak
\[
  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u).
\]
cannot reduce to the same term, since the term substitution $[\idS]\,u$ is merely a constructor.

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
\end{proof}

@@

For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are strict.
we do not need any transport since $[;]_T$ and $[]\Pi$ are strict.
However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositionally but not strictly equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
